A Formal Verification Framework for Smart Contracts Based on Symbolic Execution

### 2024-01-05 papers/static-checker/ecoop2024/ecoop2024.tex
section 2, 3, 4

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
\usepackage{microtype}%if unwanted, comment out or use option "draft"
\pdfoutput=1
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathpartir}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{comment}
\usepackage{todonotes}
\newcommand\pt{\todo[author=PT,inline]}
\lstdefinelanguage{michelson}{
  basicstyle=\fontsize{8}{9.6}\selectfont,
  morekeywords={parameter,storage,or,unit,mutez,pair,bool,address}, sensitive=false,
  morecomment=[l]{\#},
  morecomment=[\STACK]{/*}{*/},
  morestring=[b]",
}
\lstset{
  language=Caml,
  captionpos=b,
  aboveskip=-\smallskipamount,
  belowskip=-\smallskipamount,
  belowcaptionskip=0pt,
  basicstyle=\fontsize{8}{9.6}\selectfont,
  morekeywords={val}
}

@@

Michelson is a stack-based language.
A Michelson program specifies the type of its storage, the type of its single
parameter, and a list of instructions for the underlying stack machine.
A program starts on a stack with one element which contains a pair of
the current storage value (taken from the blockchain) and the parameter value, as specified by the
contract invocation.
When it finishes, the stack must contain a single element with a pair
of the final storage value (to be saved on the blockchain) and a list
of instructions.
These instructions may be token transfers or contract invocations,
among other possibilities.

Besides the standard datatypes (e.g., numbers, lists, pairs, sums)
with the expected operations,
Michelson supports a number of types and operations specific to blockchain applications (e.g.,
\texttt{mutez} for tokens, \texttt{address} for blockchain addresses).

Here is a simple example of a Michelson program:

TO BE COMPLETED

\pt{Introduce entrypoints}

\pt{I don't think the next paragraph is strictly needed}
Michelson, as a stack-based language, may pose a challenge for
developers less familiar with it compared to more mainstream languages
like Python, Java, or C. To address this, several high-level languages
such as LIGO, SmartPy, and Liquidity have been created. These
languages provide a more readable syntax, resembling popular
programming languages, allowing developers to write Tezos smart
contracts more comfortably. These high-level contracts are then
compiled down to Michelson.

@@

% For Tezos users who may lack expertise in Michelson, there is a desire
% to verify smart contract code before interaction.
Our framework contains a domain-specific language (DSL) that
facilitates the specification of smart contract properties in a
mathematical and readable format.
The DSL is designed to allow users to write formulas that align with
the capabilities of the underlying SMT solver (Z3).
The goal is to create a language
that is accessible, easily understandable even for users who are
not Michelson experts, and guarantees formulas that the solver can handle effectively.

@@

\begin{figure}[tp]
\begin{align*}
\textbf{c} & ::= 
    \text{Balance}
   \Mid \text{post\_Balance}
   \Mid \text{Amount}
   \Mid \text{Sender}
   \Mid \text{Source} 
   \Mid \text{Now}
   \Mid \text{Level} \\
   &\Mid \text{Chain\_id}
   \Mid \text{Self\_address}
   \Mid \text{Total\_voting\_power}
   \Mid \text{Voting\_power} \\
\textbf{e} & :: =  \textbf{i} \Mid  \textbf{s}  \Mid  \textbf{b} \Mid \textbf{c} \Mid (\text{x} : \text{T}) \\
&\Mid \text{Unit}  \Mid \text{Never}  \Mid \text{True}  \Mid \text{False} \\
&\Mid \text{Pair } \textbf{e } \textbf{e}  \Mid \text{Left } \textbf{e}  \Mid \text{Right } \textbf{e}  \Mid \text{Some } \textbf{e} \text{: T} \Mid \text{None T}  \\
& \Mid \textbf{e} \text{ + } \textbf{e}\Mid \textbf{e} \text{ - } \textbf{e}  \Mid \textbf{e} \text{ * } \textbf{e} \Mid \textbf{e} \text{ \/ } \textbf{e} \Mid \textbf{e} \text{ \% } \textbf{e} \Mid \textbf{e} \text{ rem } \textbf{e} \\
& \Mid \text{transfer\_token} (\textbf{e}, \textbf{e}, \textbf{e}) 
 \Mid \text{get\_map} (\textbf{e}, \textbf{e}) 
 \Mid \text{get\_big\_map} (\textbf{e}, \textbf{e}) \\
& \Mid \text{get\_contract} (\textbf{e}, \textbf{e})  
 \Mid \text{abs} (\textbf{e})  
 \Mid \text{len} (\textbf{e})
 \Mid \text{size} (\textbf{e})
\Mid \textbf{e} \text{$\wedge$} \textbf{e}  \\
\textbf{p} & :: = \textbf{e} \text{ = } \textbf{e} 
\Mid \textbf{e} \text{ != } \textbf{e}
\Mid \textbf{e} \text{ > } \textbf{e}
\Mid \textbf{e} \text{ >= } \textbf{e}
\Mid \textbf{e} \text{ < } \textbf{e}
\Mid \textbf{e} \text{ <= } \textbf{e} \\
&\Mid \text{forall} (\textbf{e}, \textbf{p})
                                            \Mid \text{exist} (\textbf{e}, \textbf{p})
                                            \Mid \textbf{p} \text{ and } \textbf{p}\Mid \textbf{p} \text{ or } \textbf{p}  \Mid \textbf{p} \text{ nor } \textbf{p} \Mid  \text{ not } \textbf{p} \Mid \textbf{p} \text{ imply } \textbf{p} \\
\end{align*}
\caption{Syntax of predicates, p}
\label{fig:syntax-predicates}
\end{figure}

\begin{lstlisting}[float,captionpos=b,caption={Smart contract property specification},label={lst:smart-contract-property-specification},numbers=left]
mcontract Name = spec
  code := {...}
  input := t
  output := t'
  pre-condition := cd
  post-condition := cd'
\end{lstlisting}
Listing \ref{lst:smart-contract-property-specification} shows the
syntax of our property specification language. Here, \KMCONTRACT,
\KWITH, \KCODE, \KINPUT, \KOUTPUT, \KPRECONDI, and \KPOSTCONDI\ are
keywords. The \KNAME\ is replaced by the contract name. In the input
and output fields, the input and output patterns are specified. Since
Michelson operates on a stack, we assume that the stack has only one
element, as specified in the input pattern. Similarly, the output
stack is assumed to be a stack with only one element, which is
specified in the output field. Thus, the syntax allows us to specify
incomplete contracts, too. For a complete smart contract, the output
should be a pair consisting of an operation list and a storage.

@@

The input and output patterns represent Michelson terms, which can be
either concrete values or symbolic values, including variables. When a
variable appears for the first time, it needs to be accompanied by its
corresponding type declaration. Subsequent mentions can omit the type
declaration.  Formally, the pre- and postcondition instances are
formulas in first-order logic that can be handled by an SMT solver (in
our case, Z3). The specification language is constituted of predicates
defined using the syntax defined in Figure
\ref{fig:syntax-predicates}. Formulas that are used in assertions or
added to solvers are terms of sort Boolean. Otherwise, terms of
Boolean and non-Boolean sort may be mixed in any combination where
sorts match up. Universal and existential quantifiers bind variables
to the scope of the quantified formula. 
\begin{lstlisting}[float,captionpos=b,caption={Specification of Add contract},label={lst:add-contract-specification},numbers=left]
mcontract Add = spec
  code := {UNPAIR; COMPARE; GE; 
          IF {ADD}{PUSH string 'Unexpected Pair'; FAILWITH}}
  input := Pair (x : int) (y : int)
  output := (z : int)
  pre-condition := (x >= 5) | (y > 3)
  post-condition := (z >= 8)
\end{lstlisting}

Let us consider the contract \texttt{Add}  in Listing
\ref{lst:add-contract-specification} as an   example of a property
specification.  Here, the program takes as input a stack with only one
element, which is a pair of two integers, symbolically specified in
the \lstinline|input| field. The program unpairs the input pair to
obtain two elements as two integers \lstinline/x/ and \lstinline/y/.
Next, the program performs a comparison between
these two integers. If \lstinline/x/ is greater than or equal to
\lstinline/y/, the program constructs a new stack with a solitary
element at the top representing the sum of  \lstinline/x/ and
\lstinline/y/, designated as \lstinline/z/ by the \lstinline|output| directive. Otherwise, the execution
terminates with an error message containing the content
\lstinline|'Unexpected pair'|. The specification continues by defining
pre- and postconditions, where the precondition states the assumption
that the integer \lstinline/x/ is greater than or equal to 5 and \lstinline/y/
is greater than or equal to 3. The symbol  \lstinline/|/ stands for
logical disjunction. The postcondition asserts that the result
\lstinline/z/ is greater than or equal to 8.

@@

\pt{can we put the source code of the contract somewhere, e.g. in an appendix?}
To illustrate entrypoint relations, let us consider a smart contract
that models an online auction. The contract has two entrypoints,
\lstinline|%bid| and \lstinline|%close|, which serve as functions for
bidding and closing the contract, respectively.  The smart contract
storage is a pair containing as its first element a boolean value
\lstinline|auction_open| that indicates whether the contract is still
open for bidding or already closed. The second element is another
nested pair, where the first element stores the highest bidder's
address, and the second element is the contract owner's address. The
current highest bid is reflected in the balance of the contract.

@@

@@

>>>>

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
\usepackage{microtype}%if unwanted, comment out or use option "draft"
\pdfoutput=1
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathpartir}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{comment}
\usepackage{todonotes}
\newcommand\pt{\todo[author=PT,inline]}
\newcommand\atha{\todo[author={@Ha},inline]}
\lstdefinelanguage{michelson}{
  basicstyle=\fontsize{8}{9.6}\selectfont,
  morekeywords={parameter,storage,or,unit,mutez,pair,bool,address}, sensitive=false,
  morecomment=[l]{\#},
  morecomment=[\STACK]{/*}{*/},
  morestring=[b]",
}
\lstset{
  language=Caml,
  captionpos=b,
  aboveskip=-\smallskipamount,
  belowskip=-\smallskipamount,
  belowcaptionskip=0pt,
  basicstyle=\fontsize{8}{9.6}\selectfont,
  morekeywords={val}
}

@@

Michelson is a stack-based, statically typed functional programming language.
A Michelson program is given by the type $S$ of its storage, the type $P$ of its single
parameter, and a typed list of instructions for the underlying stack machine.
A program starts on a stack with one element which contains a pair of
the current storage value (taken from the blockchain) and the parameter value, as specified by the
contract invocation.
When it finishes, the stack must contain a single element with a pair
of the final storage value (to be saved on the blockchain) and a list
of operations.
These operations are syntactic commands like token transfers or
contract invocations, among other possibilities.
The program thus computes a function $P\times S \to O^* \times S$
where $O$ is the type of operations.

Besides the standard datatypes (e.g., numbers, lists, pairs, sums, functions)
with the expected operations,
Michelson supports a number of types and operations specific to blockchain applications (e.g.,
\texttt{mutez} for tokens, \texttt{address} for blockchain
addresses). The type system in simple in the sense that it does not
support quantification over types.

Listing~\ref {lst:simple-program-add} contains a simple example of a Michelson program.
\begin{lstlisting}[float=tp,captionpos=b,caption={A simple Michelson program},label={lst:simple-program-add},language=michelson]
parameter int;
storage   int;
code  { 
(*                 pair int int :: [] *)
  DUP; UNPAIR;
(*   int :: int :: pair int int :: [] *)
  COMPARE; GE; 
(*         bool :: pair int int :: [] *)
  IF {UNPAIR; ADD}{PUSH string 'Unexpected Pair'; FAILWITH}
      }
\end{lstlisting}
\atha{Explain the example program}
\atha{Explain some important instructions for reference later on:
  AMOUNT, BALANCE, FAILWITH, \dots}
\atha{Introduce entrypoints}

\atha{Add citations/references to the high-level languages.}
While contracts may be implemented directly in Michelson, it can also
be used as a compilation target. Indeed, several high-level languages
such as LIGO, SmartPy, and Liquidity have been created. These
languages provide a more readable syntax, resembling popular
programming languages, allowing developers to write Tezos smart
contracts more comfortably. Our tools are applicable to all of them
because they compile to Michelson.

@@

% For Tezos users who may lack expertise in Michelson, there is a desire
% to verify smart contract code before interaction.
Our framework contains a domain-specific language (DSL) to specify
mathematical properties of smart contracts.
The DSL is designed to allow users to write formulas that align with
the capabilities of the underlying SMT solver (Z3).
The goal is to create a language
that is accessible and understandable for users who are
not Michelson experts, and guarantees formulas that the solver can handle effectively.

@@

\begin{figure}[tp]
\begin{align*}
\textbf{c} & ::= 
    \text{Balance}
   \Mid \text{post\_Balance}
   \Mid \text{Amount}
   \Mid \text{Sender}
   \Mid \text{Source} 
   \Mid \text{Now}
   \Mid \text{Level} \\
   &\Mid \text{Chain\_id}
   \Mid \text{Self\_address}
   \Mid \text{Total\_voting\_power}
   \Mid \text{Voting\_power} \\
\textbf{e} & :: =  \textbf{i} \Mid  \textbf{s}  \Mid  \textbf{b} \Mid \textbf{c} \Mid (\text{x} : \text{T}) \\
&\Mid \text{Unit}  \Mid \text{Never}  \Mid \text{True}  \Mid \text{False} \\
&\Mid \text{Pair } \textbf{e } \textbf{e}  \Mid \text{Left } \textbf{e}  \Mid \text{Right } \textbf{e}  \Mid \text{Some } \textbf{e} \Mid \text{None T}  \\
& \Mid \textbf{e} \text{ + } \textbf{e}\Mid \textbf{e} \text{ - } \textbf{e}  \Mid \textbf{e} \text{ * } \textbf{e} \Mid \textbf{e} \text{ \/ } \textbf{e} \Mid \textbf{e} \text{ \% } \textbf{e} \Mid \textbf{e} \text{ rem } \textbf{e} \\
& \Mid \text{transfer\_token} (\textbf{e}, \textbf{e}, \textbf{e}) 
 \Mid \text{get\_map} (\textbf{e}, \textbf{e}) 
 \Mid \text{get\_big\_map} (\textbf{e}, \textbf{e}) \\
& \Mid \text{get\_contract} (\textbf{e}, \textbf{e})  
 \Mid \text{abs} (\textbf{e})  
 \Mid \text{len} (\textbf{e})
 \Mid \text{size} (\textbf{e})
\Mid \textbf{e} \text{$\wedge$} \textbf{e}  \\
\textbf{p} & :: = \textbf{e} \text{ = } \textbf{e} 
\Mid \textbf{e} \text{ != } \textbf{e}
\Mid \textbf{e} \text{ > } \textbf{e}
\Mid \textbf{e} \text{ >= } \textbf{e}
\Mid \textbf{e} \text{ < } \textbf{e}
\Mid \textbf{e} \text{ <= } \textbf{e} \\
&\Mid \text{forall} (\textbf{e}, \textbf{p})
                                            \Mid \text{exist} (\textbf{e}, \textbf{p})
                                            \Mid \textbf{p} \text{ and } \textbf{p}\Mid \textbf{p} \text{ or } \textbf{p}  \Mid \textbf{p} \text{ nor } \textbf{p} \Mid  \text{ not } \textbf{p} \Mid \textbf{p} \text{ imply } \textbf{p} \\
\end{align*}
\caption{Syntax of predicates, p}
\label{fig:syntax-predicates}
\end{figure}

\begin{lstlisting}[float=tp,captionpos=b,caption={Smart contract property specification},label={lst:smart-contract-property-specification},numbers=left]
mcontract Name = spec
  code := {...}
  input := t
  output := t'
  pre-condition := cd
  post-condition := cd'
\end{lstlisting}

\subsection{Simple contracts}
\label{sec:simple-contracts}

Listing~\ref{lst:smart-contract-property-specification} shows the
syntax of our property specification language. Here, \KMCONTRACT,
\KWITH, \KCODE, \KINPUT, \KOUTPUT, \KPRECONDI, and \KPOSTCONDI\ are
keywords. The \KNAME\ is replaced by the contract name. In the input
and output fields, the input and output patterns are specified.
\atha{What patterns are supported? Patterns are not explained!}
Since
Michelson operates on a stack, we assume that the stack has only one
element, as specified in the input pattern. Similarly, the output
stack is assumed to be a stack with only one element, which is
specified in the output field. Thus, the syntax allows us to specify
incomplete contracts, too.
\pt{I disagree (though it would be a useful feature). To this end, there should be patterns to cover stacks
  with more than one element. As it stands, the input pattern can only
speak about the top of the stack and the output pattern only about the
second component of the top of the stack, which is assumed to be a
pair.\\
Moreover, a specification might want to say something about the list
of operations, too, e.g., that tokens are transferred to some address.}
For a complete smart contract, the output
should be a pair consisting of an operation list and a storage.

@@

The input and output patterns represent Michelson terms, which can be
either concrete values or symbolic values, including variables. When a
variable appears for the first time, it needs to be accompanied by its
corresponding type declaration. Subsequent mentions can omit the type
declaration.  Formally, the pre- and postcondition instances are
formulas in first-order logic that can be handled by an SMT solver (in
our case, Z3). The specification language is constituted of predicates
defined using the syntax defined in Figure~\ref{fig:syntax-predicates}. Formulas that are used in assertions or
added to solvers are terms of sort Boolean. Otherwise, terms of
Boolean and non-Boolean sort may be mixed in any combination where
sorts match up. Universal and existential quantifiers bind variables
to the scope of the quantified formula.
\atha{More explanation is needed about expressions and formulas. Why
  do you support left and right, but not lists? Why Never? What is the
meaning of transfer\_token, get\_map, get\_contract? The constants
must be explained (though you could do that in
Section~\ref{sec:background}). (Less common items could be omitted!)}

\begin{lstlisting}[float,captionpos=b,caption={Specification of Add contract},label={lst:add-contract-specification},numbers=left]
mcontract Add = spec
  code := {UNPAIR; COMPARE; GE; 
          IF {ADD}{PUSH string 'Unexpected Pair'; FAILWITH}}
  input := Pair (x : int) (y : int)
  output := (z : int)
  pre-condition := (x >= 5) | (y > 3)
  post-condition := (z >= 8)
\end{lstlisting}

Let us consider the contract \texttt{Add}  in Listing
\ref{lst:add-contract-specification} as an   example of a property
specification.  Here, the program takes as input a stack with only one
element, which is a pair of two integers, symbolically specified in
the \lstinline|input| field. The program unpairs the input pair to
obtain two elements as two integers \lstinline/x/ and \lstinline/y/.
Next, the program performs a comparison between
these two integers. If \lstinline/x/ is greater than or equal to
\lstinline/y/, the program constructs a new stack with a solitary
element at the top representing the sum of  \lstinline/x/ and
\lstinline/y/, designated as \lstinline/z/ by the \lstinline|output| directive. Otherwise, the execution
terminates with an error message containing the content
\lstinline|'Unexpected pair'|.
\atha{This Michelson code with its explanation might be an example for Section~\ref{sec:background}}
The specification continues by defining
pre- and postconditions, where the precondition states the assumption
that the integer \lstinline/x/ is greater than or equal to 5 and \lstinline/y/
is greater than or equal to 3. The symbol  \lstinline/|/ stands for
logical disjunction. The postcondition asserts that the result
\lstinline/z/ is greater than or equal to 8.
\atha{This specification is meaningless. Why not use $x\ge y$ as
  precondition and $z = x+y$ as postcondition?}

\subsection{More than one entrypoint}
\label{sec:more-than-one}

@@

\atha{can we put the source code of the contract somewhere, e.g. in an appendix?}
To illustrate entrypoint relations, let us consider a smart contract
that models an online auction. The contract has two entrypoints,
\lstinline|%bid| and \lstinline|%close|, which serve as functions for
bidding and closing the contract, respectively.  The smart contract
storage is a pair containing as its first element a boolean value
\lstinline|auction_open| that indicates whether the contract is still
open for bidding or already closed. The second element is another
nested pair, where the first element stores the highest bidder's
address, and the second element is the contract owner's address. The
current highest bid is reflected in the balance of the contract.

@@

\subsection{Entrypoint relations}
\label{sec:entrypoint-relations}

@@

\section{Static Checker}
\label{sec:static-checker-we}

We verify the contract against its specifications in the contract
module by symbolic execution. Symbolic execution involves calculating
a symbolic stack at each transition from one Michelson instruction to
the next. The symbolic interpreter is fully typed and rejects
ill-typed Michelson programs, provided the ill-typed part is reachable
from the default entry point.  The initial stack is derived from the
input pattern for each the entry point. To maintain symbolic values as
concrete as possible, some instructions may be deemed unreachable or
only reachable under specific conditions. This process yields, for
each entry point, a set of final symbolic stacks accompanied by a path
condition indicating when this final state is reachable. A state is
discarded from the final state space if the path condition is not
satisfiable, indicating that the path is unreachable. Additionally,
for each \FAILWITH\ instruction, we obtain a symbolic value for the
reported message and a path condition indicating the reachability of
this instruction. We employ the SMT solver Z3  to assess the
satisfiability of a path condition and to check smart contract
properies specified by pre- and post conditions.

The architecture of the static checker is illustrated in
Figure~\ref{fig:architecture-of-static-checker}. The system comprises
three phases. The first phase is user specification, where users
provide the Michelson code of the contract and define the properties
that the smart contract should satisfy. Subsequently, the symbolic
interpreter is run on the code with the provided input pattern.  The
path conditions are then fed into a Z3 converter, which transforms
symbolic terms into Z3 formulas. Concurrently, the property
specification, specifically pre- and post conditions, also undergoes
conversion to Z3 formulas. The results of the symbolic execution and
Z3 formulas are then input into the static checker, which verifies
whether these properties hold for the smart contract
implementation. This constitutes the verification phase.

\subsection{User specification}
\label{sec:user-specification}

In the user specification phase, users compose the specification in
our domain-specific language using the syntax outlined in
Section~\ref{sec:domain-specific-language}. Michelson code, whether
composed by users, obtained from the Tezos blockchain, or compiled
from higher-level languages, can be seamlessly used within our
framework. The specification is subsequently parsed into an Abstract
Syntax Tree (AST) in OCaml. The code and the input pattern are then
input to the symbolic interpreter. In the next phase, the pre- and
post-predicates are transformed into Z3 formulas by the Z3 converter.

\subsection{Michelson interpreter and Z3 converter}
\label{sec:mich-interpr-z3}
The symbolic interpreter implements the execution model described in
Section~\ref{sec:symbolic-execution-model}. When provided with a
concrete input, the interpreter produces a concrete value. In the
presence of symbolic input, the result comprises a set of symbolic
states represented as terms in OCaml. As Michelson's infrastructure is
implemented in OCaml, the conversion of Michelson types or terms to
their OCaml counterparts is straightforward. Each state contains the
output stack, which has only one element consisting of an operation
list and storage, along with the path conditions represented as
predicates in conjunctive normal form. This output is then fed into
the Z3 converter, which handles two different types of terms:
assertion terms from the AST representing the pre- and post-condition
specifications, and terms resulting from the interpreter. We utilize
the Z3 library in OCaml to construct this converter. Some adjustments
are necessary to convert these terms into Z3 formulas. For example, as
there is no natural number type (\TNAT) in Z3, it is injected to Z3's
integer type with a constraint asserting that the value is greater
than or equal to zero. Several constants, such as \CAMOUNT, \CBALANCE,
\CSENDER\ as well as cryptographic functions, such as \FSHA,
\FHASHKEY\ are predefined in the environment that works for any smart
contract specifications.

% \paragraph{Dealing with loop formalization}
% \label{sec:dealing-with-loop}
Our execution model formalizes loops symbolically using the \FOLD\ and
\FMAP\ functions (see
Section~\ref{sec:symbolic-execution-model}). This approach has the
advantage of allowing us to convert these loops into formulas
compatible with certain SMT solvers, such as Z3, which support these
functions. Specifically, Z3 provides \lstinline/seq.map fn s/, a map
function on sequence \lstinline/s/, and
\lstinline/(seq.foldl fn b s)/, a fold function with an initial value \lstinline/b/ on sequence
\lstinline/s/. These functions are supported for checking assertions
that contain them. Therefore, we can directly convert the symbolic
results of loops into formulas in Z3. 
\subsection{Static checker}
\label{sec:static-checker}
The static checker now performs the verification process by employing symbolic execution against the property specification in the DSL. It utilizes the results from the previous phases, including symbolic execution, Z3 formulas representing pre and post conditions, and path conditions.

We first build the input state from the input pattern and run the
symbolic interpreter on the input stack with the provided code to
generate all symbolic execution states. Symbolic execution ensures
that all states are reachable from the initial state, as any
unreachable states are discarded from the final states as their path
conditions are unsatisfiable. Using the output pattern and the pre-
and post-conditions converted by the Z3 conveter, the tool then
performs different checks. The static checker offers three distinct
features: output check, fail condition, and property check.

\subsubsection{Output check}
\label{sec:output-check}
The first check is the output check, where the checker verifies
whether the output pattern matches. The final state contains a stack
with only one element as \Stack\ \STACKCONCAT\ \EMPTYSTACK. There are
two different forms of the final state. In the case where the code is
given as incomplete smart contract code, the element \Stack\ contains
just a normal term and its corresponding type as (\Term,
\TY). However, in the case of a complete smart contract code, it has
the form (\PAIR\ \VOPERATIONLIST\ \VSTORAGE, \TPAIR\ (\TOPERATIONLIST)
\TYS) \STACKCONCAT\ \EMPTYSTACK, where \VSTORAGE\ is the term
representing the storage and \VOPERATIONLIST\ repesents a operation
list. Given an output pattern $ \tau$, the checker scans all the final
states, which do not contain a failwith stack and checks whether the
output pattern matches. In the former case, it matches $ \tau$  with
the term \Term. In the latter case, it essentially matches with the
corresponding part \VSTORAGE\ in the final states resulting from the
symbolic interpreter. When the input pattern is a concrete value, the
result from the interpreter is also concrete, and then the checker can
be used to verify if the output aligns with the expectations
specified in the output field.

Listing~\ref{lst:add-contract-specification} contains an example of an
incomplete smart contract code. The specification states that the
output of the code must be a stack that contains only one element of
type \TINT\, symbolically represented as a variable \VZ\ of type
\TINT.  In this example, there are two final states, the first one
being a failwith state recording the failwith condition when the first
element of the pair is less than the second element. The failwith
state contains the message \lstinline|'Unexpected Pair'|, which
describes the reason for the failure. The second state is a reachable
state that contains, in the top element, an integer value, which is
the sum of the two elements of the pair. The output check succeeds for
this example as expected. However, this test could fail if, for
example, the output pattern is incorrectly specified as a variable of
a type that does not match with the actual result. If the input
pattern is concrete, for instance, a pair of numbers as in
\lstinline|Pair 5 3|, then the expected output should be the number
\lstinline|8|. To check this,  the output pattern may be specified as
a concrete value.

The second example in Listing \ref{lst:auction-contract-specification}
is the specification for a smart contract code. The output pattern
represents the updated storage after executing the code. There are
several final states resulting from the symbolic execution. The
failwith state is ignored for the output check, and all other
reachable states are matched with the output term. The output check
succeeds for this example.

\subsubsection{Fail condition}
\label{sec:fail-condition}
A Michelson contract can fail due to a programmer-induced condition
caused by the instruction \FAILWITH. It terminates contract execution
with an error message reported back to the caller. This error message
includes the top value on the stack. These failures are important in
smart contract programming to guarantee that a smart contract call
should not be successful under specific circumstances. Smart contract
calls are expected to fail due to malicious behavior, such as an
unauthorized caller or invalid arguments passed along the call. For
example, the auction contract should not be closed by someone who is
not the owner of the contract, and the bidding should not succeed if
the bid is not strictly higher than the current highest bid. Several
smart contract standards, such as
\cite{@ha-supply-a-smart-contract-standard}, require the contract to fail
with specific messages under certain conditions. Therefore, it is
essential to check the fail conditions of a contract.

The ability to generate fail conditions is a potent feature of our
static checker, particularly in verifying access control properties
across various smart contracts. Specifically, in numerous smart
contract configurations, certain entrypoints are designated for
invocation only by authorized entities, for instance, by the contract
owner. Consequently, these entrypoints must result in failure if the
callers lack the necessary authorization. This crucial access control
property is verified by examining the fail conditions of a smart
contract, ensuring that unauthorized callers trigger the intended
failures in the system.

The auction contract exhibits several fail conditions. For the
entrypoint \lstinline/bid/, the fail condition is as shown in
Listing~\ref{lst:fail-condition-auction-contract-bid} and for the
\lstinline/close/ entrypoint, it is recorded in
Listing~\ref{lst:fail-condition-auction-contract-close}. For the
\lstinline/bid/ entrypoint, there are three fail conditions.  (1) If
the auction is closed, (2) if the auction is open, but the amount less
than or equal to the balance, (3) if the auction is open, the amount
is greater than the balance, but the highest bidder's address is
invalid. The last failure can happen because there are valid Michelson
addresses, which are not associated with any real account on the Tezos
blockchain, or the parameter's type is not correct: the transfer can
only succeed if the address is associated with an account and the
parameter supplied matches with the parameter expected by the Tezos
blockchain. Bidding returns the previous highest bid as part of the
transaction where the highest bidder is stored as a value of type
address. However, to receive a token transfer, this address must be
cast (by the Michelson implementation of the contract) to an implicit
contract of Michelson type (\TCONTRACT\ \TUNIT). This cast is
unavoidable, as a value of contract type cannot be stored.  However,
the cast may fail for the reasons just discussed, and its failure
leads to an error condition that is reported with a \FAILWITH\
instruction in the implementation. In our framework, this cast failure
shows up  as a fail condition, but the user's preconditions assume
that casting is successful. The \lstinline/close/ entrypoint has
similar fail conditions when the auction is closed or the owner's
address is cannot be cast. The other fail condition occurs when the
auction is open, but the sender who calls this entrypoint is not the
owner of the contract, which was stored in the contract owner part.
\atha{This discussion suggests that the specification is incomplete:
  it should contain the assertion that the highest bidder as well as
  the contract owner field are valid and point to a unit contract!}

\begin{lstlisting}[float=tp,captionpos=b,caption={Fail condition for auction contract bid entrypoint },label={lst:fail-condition-auction-contract-bid},numbers=left]
1. not (auction_open): 'Already closed'
2. auction_open && (Amount <= Balance):
   'Not higher than the current highest bid'
3. auction_open && not (Amount <= Balance) && 
   get_contract (Unit, highest_bidder) = None:
   'Invalid address' 
\end{lstlisting}

\begin{lstlisting}[float=tp,captionpos=b,caption={Fail condition for auction contract close entrypoint},label={lst:fail-condition-auction-contract-close},numbers=left]
1. not (auction_open): 'Already closed'
2. auction_open && Sender != contract_owner): 'Not owner'
3. auction_open && (Sender = contract_owner) &&
   get_contract (Unit, contract_owner) = None:
  'Invalid address'
\end{lstlisting}

\subsubsection{Property check}
\label{sec:property-check}

The static checker enables users to verify smart contract properties
by providing pre- and post-condition specifications. A property is
specified with the clause:  \lstinline/pre-condition := &&/
$\Phi_{i}$ followed by \lstinline/post-condition := &&/ $\Psi_{j}$. This
specification asserts that if all preconditions $\Phi_{i}$ hold before
smart contract execution, then all $\Psi_{j}$ must hold after
successful (i.e., without failure) termination. To verify this
property, we explore all the paths $\Pi$ 
with the path condition $\Theta$ such that ($\bigwedge \Phi_{i}$
$\land$ $\Theta$) is satisfiable. We then check that for every path
$\Pi$, $\neg$ ($\bigwedge \Phi_{i} \rightarrow \bigwedge \Psi_{j}$) is
unsatisfiable in the context given by the input for the entry point
and its corresponding final state and path condition.
\atha{Don't you have to include $\Theta$ in the check for unsat?}

Consider the specification of the Add contract in Listing
\ref{lst:add-contract-specification}. In this context, the two
variables \lstinline/x/ and \lstinline/y/ are integers. The results of
the symbolic execution are either a failwith state with the path
condition \lstinline/x/$<$\lstinline/y/,  or a state in which the
final stack has only one element, which is a symbolic variable
\lstinline/sbvar/ of type integer, with the path condition stating
that the variable \lstinline/x/ is greater than or equal to
\lstinline/y/ and the symbolic variable \lstinline/sbvar/ is equal to
the sum of \lstinline/x/ and \lstinline/y/.  For this specification,
the failwith branch is discarded because the precondition is not
satisfy with tha path condition, and then \lstinline/z/ is matched to
the symbolic variable \lstinline/sbvar/ in the other branch. In this
example, the checker then verifies that given the with the context
asserting that (\lstinline/x/ >= \lstinline/y/) $\land$ (\lstinline/z/
= \lstinline/x/ + \lstinline/y/), $\neg$ ((\lstinline/x/ >= 5) $\land$
(\lstinline/y/ > 3) $\rightarrow$ (\lstinline/z/ >= 8)) is
unsatisfiable.
\atha{I don't understand the last sentence/formula. Where does the
  $z\ge8$ come from?}

\subsubsection{Life-cycle check}
\label{sec:life-cycle-check}
\atha{To be completed}

\section{Case Studies}
\label{sec:case-stud-subs}

This section reports two case studies that we conducted with the
implemented system. Both case studies are real contracts deployed on
the Tezos blockchain and are taken from the realm of financial
applications.

\subsection{USDtz}
\label{sec:usdtz}

In a blockchain ecosystem, tokens represent digital assets that can be
transferred between accounts. These tokens are securely stored and
verified within the blockchain network. There are various types of
tokens, including native tokens (cryptocurrencies) like Bitcoin,
Ether, Tezos, and others. Additionally, tokens can be created as
overlays through smart contracts on an existing blockchain. Two common
categories are: (1) Fungible Tokens (FT), which are interchangeable
and (2) Non-Fungible Tokens (NFTs), which represent unique digital
assets, and each unit is distinct. Stablecoins, falling under fungible
tokens, are designed to maintain a stable value by tracking the value
of a fiat currency like USD or EUR.
\atha{add citations!}

While designers can create tokens that behave in any way they want, it
is advisable to implement tokens that adhere to established standards.
This adherence guarantees interoperability and compatibility across
different platforms. A token standard defines a set of rules and an
interface that smart contracts must follow to ensure compatibility
with decentralized exchanges, wallets, and other tools within the
ecosystem. For example, ERC-20 is a token standard used in Ethereum.

\atha{provide references for the standards!}
The Tezos blockchain has its own set of standards known as Financial
Applications (FA). Two notable standards are: (1) FA 1.2, designed for
fungible tokens, and (2) FA 2, a more versatile standard accommodating
both fungible and non-fungible tokens (NFTs). The standard outlines a
smart contract that functions as a ledger, associating identities with
balances. This ledger is designed to facilitate token transfer
operations and enable approvals for spending tokens from other
accounts. Any contract aiming to implement the standard must
incorporate the functions specified in
Listing~\ref{lst:fa12-interface} such as token transfer (transactions
between users), approval (allowing one user to send tokens from
another user's address),  allowance (browsing the amount of tokens
available for an account to sending from another address), user
balance (checking the balance of a user), and current supply (checking
the total supply of tokens across all addresses), where
\lstinline/view a r = pair a (contract r)/ implies an view entrypoint.
A view is an entrypoint that takes an
argument of type \lstinline/a/ and a callback that results in a value
of type \lstinline/r/. The callback in this case will be another
contract address. Further elaboration on the details of FA 1.2 can be
found in \cite{financial-applications-1.2}, and we will delve into
these functions in subsequent 
sections.
\atha{is that correct: ``callback that results in a value of type r''?
I would have expected that the callback \textbf{expects} a value of
type r.}
